---
title: "Admissible Multinomial Trial - IUT design Example"
author: "Yalin Zhu"
date: "July 18, 2016"
output: pdf_document
---



# Multinomial Trial Design with Different Output Options

In the following example, we provide four design methods for multinomial trial: Minimax (minimize the maximum sample size), Optimal (minimize the expected sample size), Admissible (minimize the Bayesian risk) and Maxpower (maximize the exact power level). 


```{r}
source("powfun_IUT.r")
source("searchfun_IUT_adm.r")
library(clinfun)
```

## Single-stage

It should be noted that single-stage design only allows Minimax and Maxpower output methods.

```{r tidy=TRUE, warning=FALSE}
# Minimax
IUT.design(method="s1",s2.rej=18, t2.rej = 12, n=80, s2.rej.delta = 1, t2.rej.delta = 1, n.delta=1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "minimax")
# Maxpower
IUT.design(method="s1",s2.rej=18, t2.rej = 12, n=80, s2.rej.delta = 1, t2.rej.delta = 1, n.delta=1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "maxpower")
```

## Two-stage

```{r tidy=TRUE, warning=FALSE}
# Minimax
IUT.design(method="s2.sf",s1.rej = 10, t1.rej = 3, s1.acc=8, t1.acc = 5, s2.rej=18, t2.rej = 12, n1=40, n2=40, n1.delta = 1, n2.delta = 1, s1.rej.delta =1, t1.rej.delta =1, s2.rej.delta =1, t2.rej.delta =1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "minimax")
# Optimal
IUT.design(method="s2.sf",s1.rej = 10, t1.rej = 3, s1.acc=8, t1.acc = 5, s2.rej=18, t2.rej = 12, n1=40, n2=40, n1.delta = 1, n2.delta = 1,  s1.rej.delta =1, t1.rej.delta =1, s2.rej.delta =1, t2.rej.delta =1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "optimal")
# Admissible
IUT.design(method="s2.sf",s1.rej = 10, t1.rej = 3, s1.acc=8, t1.acc = 5, s2.rej=18, t2.rej = 12, n1=40, n2=40, n1.delta = 1, n2.delta = 1,  s1.rej.delta =1, t1.rej.delta =1, s2.rej.delta =1, t2.rej.delta =1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "admissible")
# Maxpower
IUT.design(method="s2.sf",s1.rej = 10, t1.rej = 3, s1.acc=8, t1.acc = 5, s2.rej=18, t2.rej = 12, n1=40, n2=40, n1.delta = 1, n2.delta = 1,  s1.rej.delta =1, t1.rej.delta =1, s2.rej.delta =1, t2.rej.delta =1, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "maxpower")
```

## Two-stage early terminate with futility only

```{r error=TRUE, tidy=TRUE, warning=FALSE, message=FALSE}
# Minimax
suppressWarnings(IUT.design(method = "s2.f", s1.acc=7, t1.acc = 5, s2.rej=17, t2.rej = 13, n1=41, n2=41, s1.acc.delta =0, t1.acc.delta =0, s2.rej.delta =0, t2.rej.delta =0, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "minimax"))  
# Optimal
suppressWarnings(IUT.design(method = "s2.f", s1.acc=7, t1.acc = 5, s2.rej=17, t2.rej = 13, n1=41, n2=41, s1.acc.delta =0, t1.acc.delta =0, s2.rej.delta =0, t2.rej.delta =0, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "optimal"))
# Admissible
suppressWarnings(IUT.design(method = "s2.f", s1.acc=7, t1.acc = 5, s2.rej=17, t2.rej = 13, n1=41, n2=41, s1.acc.delta =0, t1.acc.delta =0, s2.rej.delta =0, t2.rej.delta =0, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "admissible"))
# Maxpower
suppressWarnings(IUT.design(method = "s2.f", s1.acc=7, t1.acc = 5, s2.rej=17, t2.rej = 13, n1=41, n2=41, s1.acc.delta =0, t1.acc.delta =0, s2.rej.delta =0, t2.rej.delta =0, p0.s = 0.15, p0.t = 0.25, p1.s = 0.3, p1.t= 0.1, output = "maxpower"))
```
